---
title: "ADA Final Project"
author: "S.Francis S."
date: "2024-10-24"
output:
  pdf_document: default
  word_document:
    reference_docx: "custom-reference.docx"
  html_document:
    df_print: paged
always_allow_html: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
# TITLE. The association between receiving money or gifts from a sexual partner and condom usage among teenagers in Kenya, 2003 – 2005.

## Install packages and open libraries
```{r cars}
pacman::p_load(tidyverse, readr, haven, table1, DiagrammeR, rsvg, dagitty, ggdag, Hmisc, DiagrammeRsvg, odds.n.ends, blorr, lmtest, car, broom, jtools)

```

```{r}
setwd("C:/Users/Ssemujju Francis/Box/Franx Doxs/MPH Program/Yr 2 Semester 1/Advanced Data Analysis/Final Project")

```

```{r}
### Load dataset
sur_data <- read_dta("HIKAP_survey.dta")

# View the first few rows of the data
head(sur_data)
```

```{r}
summary(sur_data)
```

# Data management for variables to be used in this analysis
```{r}
# look at column names
colnames(sur_data)
```

# Select variables of interest
```{r}
sur_data <- sur_data %>%
  select(
    condomlast,       # Dependent variable
    giftpartner,      # Independent variable
    regpartner,       # Effect modifier
    age,              # Adjustment variable: age
    sex,              # Adjustment variable: gender
    realdivision,     # Adjustment variable: geographical division
    boysonlyschool,   # Adjustment variable: boys-only school
    girlsonlyschool   # Adjustment variable: girls-only school
  )

# Check column names
colnames(sur_data)

```


# conduct some data management
```{r}
# Conduct data management for the HIKAP dataset
sur_data <- sur_data %>%
  # Recode variables
  mutate(
    giftpartner_f = factor(
      giftpartner,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    condomlast_f = factor(
      condomlast,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    regpartner_f = factor(
      regpartner,
      levels = c(1, 2),
      labels = c("Has Partner", "No Partner")
    ),
    # Convert sex variable to factor with meaningful labels
    sex_f = factor(
      sex,
      levels = c("M", "F"),
      labels = c("Male", "Female")
    ),
    # Ensure age is numeric
    age = as.numeric(age)
  )

# Verify recoded variables
table(sur_data$giftpartner, sur_data$giftpartner_f, useNA = "always")
table(sur_data$condomlast, sur_data$condomlast_f, useNA = "always")
table(sur_data$regpartner, sur_data$regpartner_f, useNA = "always")
table(sur_data$sex, sur_data$sex_f, useNA = "always")


```
# Figure 1 
## Let's make a figure 1 that shows exclusions to arrive at your analytic dataset
## First we will make the following exclusions one at a time
- Exclusion 1: Remove records with missing data or "Unknown" values for the independent variable (giftpartner).
```{r}
# Exclusion 1: Remove records with missing or "Unknown(N/A)" values for giftpartner_f
sur_data1 <- sur_data %>%
  filter(!is.na(giftpartner_f), giftpartner_f != "N/A")

# Count the excluded records
ex1 <- nrow(sur_data) - nrow(sur_data1)
ex1

```

- Exclusion 2: Remove records with missing or "Unknown (N/A)" values for the dependent variable (condomlast).
```{r}
# Exclusion 2: Remove records with missing or "Unknown/N/A" values for condomlast_f
sur_data2 <- sur_data1 %>%
  filter(!is.na(condomlast_f), condomlast_f != "N/A")

# Count the number of exclusions
ex2 <- nrow(sur_data1) - nrow(sur_data2)
ex2

```

- Exclusion 3: Remove records with missing data on key covariates.
Include age, sex_f, realdivision, and regpartner_f:
```{r}


sur_data3 <- sur_data2 %>%
  drop_na(age, sex_f, regpartner_f, realdivision, boysonlyschool, girlsonlyschool)

ex3 <- nrow(sur_data2) - nrow(sur_data3)
ex3

```

# Exclusion 4: Handle outliers, duplicates, and invalid data
```{r}
# Exclusion 4: Handle outliers, duplicates, and invalid data
sur_data4 <- sur_data3 %>%
  # Exclude unrealistic ages
  filter(age >= 13 & age <= 19) %>%
  # Exclude invalid school type values
  filter(boysonlyschool %in% c(0, 1), girlsonlyschool %in% c(0, 1)) %>%
  # Remove duplicates
  distinct() %>%
  # Exclude inconsistent school type records
  filter(!(boysonlyschool == 1 & girlsonlyschool == 1))

# Calculate the number of exclusions for Exclusion 4
ex4 <- nrow(sur_data3) - nrow(sur_data4)
ex4

```

# Create the Flowchart With branching 
```{r}
# Generate the flowchart
grViz("digraph flowchart {
     
      node [fontname = Helvetica, shape = rectangle, fontsize=15] 
      
      # Define the nodes
      node1 [label = '@@1']
      node2 [label = '@@2']
      node3 [label = '@@3']
      node4 [label = '@@4']
      node5 [label = '@@5']
      
      # Define the edges
      node1 -> node2 -> node3 
      node3 -> node4 # branching
      node3 -> node5 # branching
}
      # Define the labels for each node
      [1]: 'Original records received for HIKAP survey n = 9,951'
      [2]: 'Excluding 6,417 records with missing or Unknown values \\n on giftpartner n = 3,534'
      [3]: 'Excluding 880 records with missing data \\n on condom usage n = 2,654'
      [4]: 'Excluding 126 records with missing \\n key covariates (age, sex, etc.) n = 2,528'
      [5]: 'Excluding 1,746 records with missing data \\n on school types n = 782'
      ")

```


# Wrapping text in second box with \\n
```{r}
# Note I attach to an object called figure 1 for exporting below
figure1 <- grViz("digraph flowchart {
     
      node [fontname = Helvetica, shape = rectangle, fontsize=15] 
      
      # Define nodes
      node1 [label = '@@1']
      node2 [label = '@@2']
      node3 [label = '@@3']
      node4 [label = '@@4']
      node5 [label = '@@5']
      
      # Define edges
      node1 -> node2 -> node3 
      node3 -> node4 # branching
      node3 -> node5 # branching
}
      # Node text with line breaks
      [1]: 'Original records n = 9,951'
      [2]: 'Excluding 6,417 records with \\n missing data and N/A on giftpartner n = 3,534'
      [3]: 'Excluding 880 records with \\n missing data on condom usage n = 2654'
      [4]: 'Excluding 126 records with \\n missing data on key covariates \\n (age, sex, division, regpartner) n = 2,528'
      [5]: 'Excluding 1,746 records with \\n missing data on survival variables n = 782'
      ")
figure1

```




# export figure
```{r}
figure1 %>%

DiagrammeRsvg::export_svg() %>% 
  charToRaw() %>% 
  rsvg::rsvg_pdf("Figure_1_HIKAP.pdf")
```

# Creating a Table 1 using the table1 function. 

# Adding labels to the variables because the raw variables are not publication-quality. This is important in producing a high-quality table.
```{r}
label(sur_data4$age) <- "Age (years)" # Age of respondent
label(sur_data4$sex_f) <- "Gender (Male/Female)" # Gender recoded as Male/Female
label(sur_data4$giftpartner_f) <- "Received Gifts or Money from Partner" # Independent variable
label(sur_data4$condomlast_f) <- "Used Condom During Last Sexual Encounter" # Dependent variable
label(sur_data4$regpartner_f) <- "Has Regular Partner" # Effect modifier
label(sur_data4$boysonlyschool) <- "Attends Boy's Only School" # School type variable
label(sur_data4$girlsonlyschool) <- "Attends Girl's Only School" # School type variable

# Generate Table 1 stratified by condomlast_f with adjusted labels and summary
table1(~ age + sex_f + giftpartner_f + regpartner_f + boysonlyschool + girlsonlyschool + realdivision| condomlast_f, 
       data = sur_data4, 
       overall = "Total", # Add overall summary for all respondents
       rowlabelhead = "Variable", # Header label for row names
       footnote = "SD = standard deviation. Data stratified by whether the respondent received gifts/money from a partner.")

```




```{r}
# Load necessary libraries
library(flextable)
library(officer)

# Convert Table 1 to a flextable object
table1_flex <- as_flextable(table1(~ age + sex_f + giftpartner_f + regpartner_f + boysonlyschool + girlsonlyschool | condomlast_f, 
                                   data = sur_data4, 
                                   overall = "Total", 
                                   rowlabelhead = "Variable", 
                                   footnote = "SD = standard deviation. Data stratified by condom usage."))

# Create a Word document
doc <- read_docx()

# Add the table to the document
doc <- body_add_flextable(doc, value = table1_flex)

# Save the Word document to your drive
print(doc, target = "Table1_CondomUsage.docx")

```


# Brief interpritation
Age: Respondents who reported using a condom during their last sexual encounter were slightly older on average (16.4 years, SD = 1.38) than those who did not (16.2 years, SD = 1.48). The median age was the same across both groups (16 years). This indicates that there is little difference in age distribution between the two groups.
Gender: A similar proportion of males reported using condoms (64.3%) compared to those who did not (64.6%). Similarly, the percentage of females was comparable between condom users (35.7%) and non-users (35.4%). This suggests that gender distribution is nearly identical between the groups and may not be a strong determinant of condom usage.

Received Gifts or Money from Partner: A slightly higher percentage of respondents who received gifts or money from a partner reported using condoms (51.4%) compared to those who did not use condoms (49.5%). Conversely, a slightly lower percentage of respondents who did not receive gifts/money used condoms (48.6%) compared to those who did not use condoms (50.5%). This suggests a weak association between receiving gifts/money and condom usage.

Has Regular Partner: Respondents with a regular partner were more likely to use condoms (46.1%) compared to those who did not have a regular partner (39.2%). Conversely, respondents without a regular partner were more likely not to use condoms (60.8%) compared to those who used condoms (53.9%). This indicates that having a regular partner might slightly influence condom usage.

Attends Boy's Only School: Respondents who used condoms had a slightly lower mean attendance at boys-only schools (0.236) compared to non-users (0.246). However, the median attendance was 0 for both groups, indicating that most respondents did not attend boys-only schools. This variable appears to have limited variability and influence on condom usage.

Attends Girl's Only School: Similarly, attendance at girls-only schools was comparable between the groups, with a mean of 0.0927 for condom users and 0.0892 for non-users. The median attendance was 0 for both groups, showing that this variable also has limited variability and impact on condom usage.

# General Observations:
Age, gender, and school type (boys-only and girls-only) show minimal differences between respondents who used condoms and those who did not.
Having a regular partner shows a slightly stronger association with condom usage, while receiving gifts/money shows a weak association.
These findings suggest that further analysis, such as multivariable logistic regression, is needed to assess the independent effects of these variables while controlling for confounders.


# Startistical tests

```{r}
# Cross-tabulation of giftpartner_f and condomlast_f
table(sur_data4$giftpartner_f, sur_data4$condomlast_f)

# Descriptive summary of continuous variables
summary(sur_data4$age)
summary(sur_data4$boysonlyschool)
summary(sur_data4$girlsonlyschool)


```

# Bivariate Analysis
- Use a Chi-Square test to assess the association between giftpartner_f and condomlast_f.
- Perform a t-test or Wilcoxon test for continuous variables like age.
```{r}
# Chi-Square test for association between giftpartner_f and condomlast_f
chisq_test <- chisq.test(table(sur_data4$giftpartner_f, sur_data4$condomlast_f))
chisq_test

```
# # Run a univariable (unadjusted) logistic regression model for receiving gifts/money predicting condom usage, with log odds and OR.
```{r}
model1 <- glm(condomlast_f ~ giftpartner_f, data = sur_data4, family = "binomial")

# View the model summary to get log odds
summary(model1) 

# Exponentiate coefficients to get Odds Ratios (ORs)
exp(model1$coefficients)

# Get 95% Confidence Intervals (CIs) for ORs
exp(confint(model1))


    odds.n.ends(model1) 

```

# Interpretation:

For individuals who did not receive gifts/money, the odds of using a condom are 7.8% higher compared to those who did receive gifts/money. However, this association is not statistically significant (p = 0.602), and the confidence interval (0.813–1.429) includes 1, suggesting that there may be no real effect.
The intercept suggests that the baseline odds of using a condom (when the predictor variable is at its reference level) are 1.15 times higher. Again, this is not significant.

# Logistic regression with giftpartner_f as a binary predictor
```{r}

model <- glm(condomlast_f ~ giftpartner_f + age + sex_f + regpartner_f + boysonlyschool + girlsonlyschool + realdivision, 
             data = sur_data4, 
             family = "binomial")

# Summary of the model
summary(model)

# Odds Ratios and Confidence Intervals
exp(model$coefficients) # ORs
exp(confint(model)) # CIs

```

# Brief interpretation:


```{r}
library(car)
vif_model <- glm(condomlast_f ~ giftpartner_f + age + sex_f + regpartner_f, 
                 family = "binomial", data = sur_data4)
vif(vif_model)

```


# Let's refine the model by removing variables with high p-value
```{r}
# Simplified logistic regression model
model_refined <- glm(
  condomlast_f ~ giftpartner_f + age + sex_f + regpartner_f, 
  data = sur_data4, 
  family = "binomial"
)

summary(model_refined)
exp(model_refined$coefficients) # ORs
exp(confint(model_refined)) # CIs

```

Create tabel for presentation of results. 
```{r}
# Prepare the table data
table_data <- data.frame(
  Variable = c("Intercept", "No Gifts (Ref: Yes)", "Age", "Female (Ref: Male)", "No Partner (Ref: Partner)"),
  Estimate = c(1.61209, 0.03550, -0.09777, -0.07618, 0.29058),
  OR = c(5.01, 1.04, 0.91, 0.93, 1.34),
  `95% CI` = c("0.97–26.36", "0.78–1.38", "0.82–1.00", "0.69–1.25", "1.00–1.79"),
  `p-value` = c(0.056, 0.808, 0.054, 0.621, 0.049)
)

# Create the table
knitr::kable(table_data, 
             col.names = c("Variable", "Estimate", "OR", "95% CI", "p-value"),
             caption = "Logistic Regression Results for Condom Use")

```

```{r}
# Logistic regression with interaction term
interaction_model <- glm(condomlast_f ~ giftpartner_f * regpartner_f + age + sex_f + boysonlyschool + girlsonlyschool + realdivision, 
                         family = "binomial", data = sur_data4)

summary(interaction_model)

# Perform LR test to assess interaction
lrtest(main_model, interaction_model)


```






